---
title: "Results — provisional"
subtitle: "Conservation distributions, BBB vs liver, and the three fixes that change them"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
knitr::opts_knit$set(root.dir = "..")
```

```{=html}
<div class="prov-banner">
  <h4>These findings are provisional pending three fixes</h4>
  <p>Steps <strong>1</strong>, <strong>4c</strong> and <strong>5d</strong> feed everything below.
     Until they are corrected and steps 5c&ndash;5h re-run, no figure on this page should be
     cited. Each finding carries its own verdict underneath. One comparison already reverses
     direction once the orthologue rule is applied symmetrically:</p>
  <div class="prov-flip">
    <div><span class="fl">As published &mdash; every match kept on the BBB side only</span><span class="fv">BBB 84.12 &middot; liver 84.32</span><span class="fd down">BBB less conserved</span></div>
    <div><span class="fl">Same rule applied to both sides</span><span class="fv">BBB 85.04 &middot; liver 84.47</span><span class="fd up">BBB more conserved</span></div>
  </div>
</div>
```

## Key figures

```{r v2-results}
df <- read_csv("processed/conservation_scores_combined_v2.csv",
               show_col_types = FALSE)

bbb   <- df[df$Gene_Set == "BBB", ]
liver <- df[df$Gene_Set == "Liver_Control", ]

results <- data.frame(
  Comparison = c("Human vs Mouse — % identity",
                  "Human vs Macaque — % identity",
                  "Human vs Mouse — dN/dS (corrected)",
                  "Human vs Macaque — dN/dS (corrected)"),
  BBB_median = c(
    round(median(bbb$PctId_Human_Mouse,      na.rm = TRUE), 2),
    round(median(bbb$PctId_Human_Macaque,    na.rm = TRUE), 2),
    round(median(bbb$dNdS_Human_Mouse_v2,    na.rm = TRUE), 3),
    round(median(bbb$dNdS_Human_Macaque_v2,  na.rm = TRUE), 3)
  ),
  Liver_median = c(
    round(median(liver$PctId_Human_Mouse,      na.rm = TRUE), 2),
    round(median(liver$PctId_Human_Macaque,    na.rm = TRUE), 2),
    round(median(liver$dNdS_Human_Mouse_v2,    na.rm = TRUE), 3),
    round(median(liver$dNdS_Human_Macaque_v2,  na.rm = TRUE), 3)
  )
)
kable(results, caption = "BBB vs liver conservation — final corrected values")
```

```{r v2-pid-plot, fig.cap="Distribution of % nucleotide identity — BBB vs Liver"}
df_long <- df |>
  select(Gene_Set, PctId_Human_Mouse, PctId_Human_Macaque) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison",
                       values_to = "PctIdentity") |>
  filter(!is.na(PctIdentity)) |>
  mutate(Comparison = sub("PctId_", "", Comparison),
         Comparison = sub("_", " vs ", Comparison))

ggplot(df_long, aes(x = PctIdentity, fill = Gene_Set)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Comparison) +
  scale_fill_manual(
    values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4"),
    labels = c("BBB" = "BBB (red)", "Liver_Control" = "Liver (blue)")
  ) +
  labs(x = "% nucleotide identity", y = "Density",
       title = "BBB vs Liver: distribution of % identity",
       fill = "Gene set",
       caption = "Purple = region where the BBB and Liver distributions overlap") +
  theme_minimal() +
  theme(plot.caption = element_text(colour = "#7b4f6e", face = "italic", size = 9))
```

```{r v2-dnds-plot, fig.cap="Distribution of dN/dS (codon-aligned)"}
df_dnds <- df |>
  select(Gene_Set, dNdS_Human_Mouse_v2, dNdS_Human_Macaque_v2) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison", values_to = "dNdS") |>
  filter(!is.na(dNdS), dNdS < 2)

ggplot(df_dnds, aes(x = dNdS, fill = Gene_Set)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Comparison) +
  scale_fill_manual(
    values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4"),
    labels = c("BBB" = "BBB (red)", "Liver_Control" = "Liver (blue)")
  ) +
  labs(x = "dN/dS (codon-aligned)", y = "Density",
       title = "BBB vs Liver: distribution of dN/dS",
       fill = "Gene set",
       caption = "Purple = region where the BBB and Liver distributions overlap") +
  theme_minimal() +
  theme(plot.caption = element_text(colour = "#7b4f6e", face = "italic", size = 9))
```

```{r pid-vs-dnds-plot, fig.cap="% identity vs dN/dS — both metrics tell the same story"}
bbb_scatter <- bbb |>
  filter(!is.na(PctId_Human_Mouse), !is.na(dNdS_Human_Mouse_v2), dNdS_Human_Mouse_v2 < 2)

ggplot(bbb_scatter, aes(x = PctId_Human_Mouse, y = dNdS_Human_Mouse_v2)) +
  geom_point(alpha = 0.25, color = "#d62728", size = 1.2) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.8) +
  labs(x = "% nucleotide identity (Human vs Mouse)", y = "dN/dS (codon-aligned)",
       title = "% identity and dN/dS are correlated but capture different signal",
       subtitle = "BBB genes only — Human vs Mouse") +
  theme_minimal()
```

```{r purifying-selection-bar, fig.cap="Proportion of genes under strong purifying selection (dN/dS < 0.3)"}
purify_df <- df |>
  filter(!is.na(dNdS_Human_Mouse_v2) | !is.na(dNdS_Human_Macaque_v2)) |>
  group_by(Gene_Set) |>
  summarise(
    `Human vs Mouse`   = mean(dNdS_Human_Mouse_v2   < 0.3, na.rm = TRUE) * 100,
    `Human vs Macaque` = mean(dNdS_Human_Macaque_v2 < 0.3, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison", values_to = "Pct")

ggplot(purify_df, aes(x = Comparison, y = Pct, fill = Gene_Set)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4")) +
  labs(x = NULL, y = "% of genes with dN/dS < 0.3",
       title = "Proportion of genes under strong purifying selection",
       subtitle = "dN/dS < 0.3 = conventional threshold for strong constraint",
       fill = "Gene set") +
  theme_minimal()
```

---

## Headline findings

1. **Macaque is the closest model to human BBB.** 96.8% nucleotide identity at the coding-sequence level. For almost every gene, macaque sits above the diagonal — more similar to human than mouse.

2. **Mouse is more divergent (84.1%) but still highly conserved.** Mouse models remain biologically defensible, but ~16% sequence divergence is present. Where mouse and human BBB phenotypes differ clinically, this divergence is one possible explanation.

3. **BBB genes are under stronger purifying selection than liver genes** — most visibly in the human-macaque comparison (BBB dN/dS 0.152 vs Liver 0.196, p < 10⁻⁶ after Bonferroni correction). The signal is weaker in human-mouse because at ~80 million years, all genes have accumulated extensive change.

4. **Gene duplication produces dramatic divergence.** Paralogous mouse genes (e.g. *Abcb1a* + *Abcb1b* for human ABCB1) are only ~47% identical to human, vs ~85% for one-to-one orthologues. Effect size is large (rank-biserial r = 0.72).

5. **Conservation and human BBB expression are independent.** A gene's validation score (0–3 datasets) does not predict how conserved it is at the sequence level.

## Biological interpretation

Strong purifying selection on BBB genes makes biological sense. The BBB controls what enters and exits the brain — broken transporters or tight junctions would let toxins in or block essential nutrients. Evolution actively removes protein-changing mutations from these genes, more aggressively than for liver genes.

The fact that this signal is most visible in the human-macaque comparison — not human-mouse — reflects evolutionary timescale. Subtle differential selection is detectable when species are close enough that not all genes have saturated their substitutions. At ~80 million years for human-mouse, this signal is lost in the noise.

---

## Findings, re-graded against the audit

The five findings above are stated as originally written. This is what the audit does to
each of them.

```{=html}
<div class="regrade shifts">
  <h4>1. Macaque is the closest model to human BBB <span class="audit-chip">Direction holds, number unreliable</span></h4>
  <p class="as-stated">As stated: 96.8% nucleotide identity; macaque sits above the diagonal for almost every gene.</p>
  <p class="verdict">% identity was computed in <code>step5c</code> and <strong>never recomputed</strong> after the codon-aware fix. <strong>224 one-to-one macaque pairs score under 85%</strong>, which is biologically impossible — <code>SLC2A1</code> comes out at 71.4%. That is the longest-transcript heuristic comparing non-corresponding isoforms. Macaque being closer to human than mouse is safe; the figure 96.8% is not.</p>
</div>

<div class="regrade shifts">
  <h4>2. Mouse is more divergent (84.1%) but still highly conserved <span class="audit-chip">Number unreliable</span></h4>
  <p class="as-stated">As stated: ~16% sequence divergence; mouse models remain biologically defensible.</p>
  <p class="verdict">The same % identity problem, plus the gene list itself is wrong. This figure describes a set that is largely <strong>pericyte and developmental-timing genes</strong>, not BBB-enriched ones. The claim is not disproven — it has not yet been tested on the right genes.</p>
</div>

<div class="regrade breaks">
  <h4>3. BBB genes are under stronger purifying selection than liver genes <span class="audit-chip">Withdrawn</span></h4>
  <p class="as-stated">As stated: BBB dN/dS 0.152 vs liver 0.196, p &lt; 10⁻⁶ after Bonferroni correction.</p>
  <p class="verdict">The comparator is not a liver control. <strong>86.6% of it is brain-expressed</strong>, with brain and liver medians effectively identical (31.3 vs 31.2). It is a housekeeping set — and housekeeping genes are unusually conserved — so the test compared BBB genes against one of the most conserved groups in the genome and read the result as tissue-specific selection. <strong>The statistics are correct; the contrast is not.</strong> This claim cannot be made until the control is rebuilt, and step 5d's asymmetric orthologue rule has to be fixed at the same time.</p>
</div>

<div class="regrade clean">
  <h4>4. Gene duplication produces dramatic divergence <span class="audit-chip">Survives</span></h4>
  <p class="as-stated">As stated: paralogous mouse genes ~47% identical vs ~85% for one-to-one orthologues; rank-biserial r = 0.72.</p>
  <p class="verdict">This is a contrast <strong>within</strong> the BBB set, so it does not depend on the liver control, and the effect is far too large to be an artefact of the % identity problem. It is the strongest surviving result in the project — worth stating plainly, because a page about mistakes that reports nothing left standing is its own kind of distortion.</p>
</div>

<div class="regrade breaks">
  <h4>5. Conservation and human BBB expression are independent <span class="audit-chip">Uninformative</span></h4>
  <p class="as-stated">As stated: a gene's 0–3 validation score does not predict how conserved it is.</p>
  <p class="verdict">The validation score barely carries information. One of its three components read <strong>only Yang's arterial sheet</strong>; another asked whether a gene was detected at all, which <strong>88.5% of genes pass</strong> — and that component alone is why 1,131 of 1,526 genes score exactly 1. A null result against a near-constant variable says nothing in either direction.</p>
</div>
```

## Limitations

### What counts as a canonical BBB gene was never pinned down

The project uses the phrase "canonical BBB genes" without ever saying which genes it means.
The working definition is the one the field uses: genes consistently identified across
multiple independent transcriptomic studies as highly expressed and functionally important in
brain endothelial cells. "Canonical" does not mean the only important BBB genes, it means the
well validated ones.

Four papers read for the [literature review](literature-review.qmd) do name them, so the
master list of 1,526 genes can be checked against those names.

| Canonical marker | On the list | Note |
|---|---|---|
| `SLC2A1` (GLUT1) | yes | glucose transporter |
| `MFSD2A` | yes | DHA transport, suppresses transcytosis |
| `ABCB1` (P-gp) | yes | main efflux pump |
| `ABCG2` (BCRP) | yes | efflux pump |
| `SLC7A5` (LAT1) | yes | amino acid transport |
| `SLCO1C1`, `SLC22A8` | yes | organic anion transport |
| `LSR`, `FLT1`, `TFRC` | yes | |
| `OCLN` (occludin) | yes | tight junction |
| `CLDN5` | **no** | tight junction, the most cited BBB gene of all |
| `TJP1` (ZO-1) | **no** | tight junction scaffold |
| `PECAM1`, `KDR`, `CDH5` | no | general endothelial markers |

: {tbl-colwidths="[30,14,56]"}

**Two different reasons for absence, which should not be conflated.** `PECAM1`, `KDR` and
`CDH5` are general endothelial markers expressed in endothelium everywhere. An
enrichment-based list is supposed to exclude them, and their absence is correct.

`CLDN5` and `TJP1` are excluded by the same rule, because they too are expressed in all
endothelia, but here the rule removes the machinery that makes the barrier a barrier. The
clearest illustration is the claudin family: **of 1,526 genes, the only claudin present is
`CLDN6`**, which is not a BBB claudin.

**The transporters are better covered.** The list contains 52 `SLC` genes, about 3.4% of it,
and 9 `ABC` transporters, including all the ones the literature names.

**What this means for every result on this page.** Enrichment and importance are not the same
thing, and this list selects for the first. Any statement here about "BBB genes" should be
read as *genes differentially enriched in brain endothelium in two mouse studies*, not *genes
that make the barrier work*. The tight junction proteins, which are the barrier, are almost
entirely missing.

**The fix** is to name the canonical set explicitly, check membership, and either state that
the enrichment rule excludes some of them or carry them as a named supplementary set that is
scored alongside. This is separate from the [three fixes](audit.qmd) and does not depend on
them.

### Methodological caveats

- **Pairwise dN/dS, not phylogenetic.** A more rigorous analysis uses PAML `codeml` on a multi-species alignment with an explicit phylogenetic tree.
- **Longest-transcript heuristic.** MANE Select / Ensembl Canonical (`step5b_alt_canonical_transcripts.R`) are more standard. The canonical FASTAs are already generated.
- **Multiple testing correction was applied to primary tests only.** Stratified subgroup analyses (by Daneman filter, validation score, paralogue type) are exploratory.

### Biological caveats

- **The starting gene list is enrichment-based.** See the section below, which sets out what that excludes.
- **Liver is one specific control choice.** A different control (brain non-BBB genes, genes matched by protein domain) could give a different picture.
- **Coding sequence only.** Regulatory regions (promoters, enhancers, UTRs) — where most cross-species expression-level differences live — were deliberately excluded.

---
