---
title: "Blood-Brain Barrier Gene Conservation Across Human, Mouse, and Macaque"
author: "Yara Ismail"
date: today
format:
  html:
    toc: false
    smooth-scroll: true
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
    output-file: index.html
  pdf:
    toc: true
execute:
  warning: false
  message: false
  cache: true
---

::: {.panel-tabset}

## Introduction

The standard preclinical pipeline for drugs targeting the brain looks like this:

> **Mouse** → validate → **Macaque** → validate → **Human clinical trial**

Each arrow is a translational leap. A drug that works in mouse must survive the jump to macaque, then to human. If the underlying molecular machinery — the blood-brain barrier (BBB) genes that control what enters and exits the brain — differs substantially between species, that pipeline is fragile. If it is conserved, the pipeline has a stronger molecular foundation.

This project asks:

> **How similar are BBB genes across all three rungs of this translational ladder — at the DNA sequence level?**

We quantify this using two complementary metrics:

- **% nucleotide identity** — how many DNA letters match between species
- **dN/dS ratio** — whether evolution is actively *protecting* these genes from protein-changing mutations (dN/dS << 1 = purifying selection)

The analysis covers three pairwise comparisons: Human vs Mouse, Human vs Macaque, and Mouse vs Macaque. BBB genes are compared against a liver control set to establish whether any conservation signal is BBB-specific or just a property of all expressed genes.

### What are "canonical BBB genes"?

The BBB is formed by brain endothelial cells (ECs) — the cells lining the brain's blood vessels. "Canonical BBB genes" refers to genes that are *consistently identified across multiple independent transcriptomic studies* as highly expressed and functionally important in these cells. This includes:

- **Tight junction proteins** — CLDN5, OCLN, TJP1 (seal gaps between cells to block free diffusion into the brain)
- **Transporters** — SLC2A1/GLUT1 (glucose), ABCB1/P-gp (drug efflux pump), SLC7A5 (amino acids)
- **Signalling receptors** specific to brain endothelium

"Canonical" does not mean "the only important BBB genes" — it means well-validated across datasets.

### Pipeline overview

```
Step 1    Build master BBB gene list (Daneman + Munji)
Step 2    Process Wälchli single-cell EC expression
Step 3    Map gene symbols → Ensembl IDs  [HomoloGene attempt — superseded]
Step 3b   Re-run mapping via BioMart + add orthology metadata
Step 4    Download GTF genome annotations  [removed — unnecessary]
Step 4b   Cross-reference against human EC datasets
Step 4c   Build liver control set from Human Protein Atlas
Step 5a   Download CDS (coding sequence) FASTAs
Step 5b   Extract one CDS per gene (longest transcript)
Step 5b*  Alternative: Ensembl Canonical / MANE Select transcripts
Step 5c   Pairwise alignment + % identity + dN/dS for BBB genes
Step 5d   Same for liver control genes
Step 5e   Statistical comparison (Wilcoxon)
Step 5f   Recompute dN/dS with codon-aware alignment  [bug fix]
Step 5g   Multiple testing correction
Step 5h   Direct mouse vs macaque comparison
```

> **Note on step numbering:** Steps jump from 3 to 3b and from 4 to 4b — these reflect the revision history. See the relevant tabs for details.

---

## Prerequisites

::: {.callout-note}
## Read-only vs reproducible

This HTML document is **read-only** — you can read the code and results but cannot run it here. To reproduce the analysis from scratch, open the scripts in RStudio (or VS Code with the R extension) and run each `scripts/step*.R` file in order.

The heavy computation (genome alignments, BioMart queries) was already run; this tutorial reads those saved outputs.
:::

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
knitr::opts_knit$set(root.dir = "..")
```

### R packages

Install all required packages before running any scripts:

```r
# CRAN packages
install.packages(c(
  "readr", "dplyr", "ggplot2", "tidyr", "knitr",
  "readxl", "homologene", "seqinr"
))

# Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "biomaRt",       # Ensembl cross-species orthologue queries
  "Biostrings",    # reading/writing FASTA files, DNA translation
  "pwalign"        # pairwise sequence alignment (Needleman-Wunsch)
))

# Seurat — required for Step 2 only
install.packages("Seurat")
```

### Raw data files

| File | Where to get it | Place it at |
|---|---|---|
| Daneman 2010 S3 | [PLOS ONE paper](https://doi.org/10.1371/journal.pone.0013741) → Supplementary S3 | `raw_data/daneman_2010/Daneman2010_S3_CoreBBBGenes_BrainEC_Enriched.xls` |
| Daneman 2010 S4 | Same paper → Supplementary S4 | `raw_data/daneman_2010/Daneman2010_S4_BrainEC_vs_LiverEC_Enriched.xls` |
| Daneman 2010 S5 | Same paper → Supplementary S5 | `raw_data/daneman_2010/Daneman2010_S5_BrainEC_vs_LungEC_Enriched.xls` |
| Munji 2019 S5 | [Nat. Neurosci. paper](https://doi.org/10.1038/s41593-019-0468-z) → Supplementary S5 | `raw_data/munji_2019/Munji2019_S5_BBBEnrichedGenes.xlsx` |
| Wälchli 2024 Seurat object | [Nature paper](https://doi.org/10.1038/s41586-024-07069-2) → repository | `raw_data/walchli_2024/Walchli2024_AdultBrain_TemporalLobe_SortedEndothelial_SeuratObject.rds.gz` |
| HPA RNA consensus | **Auto-downloaded** by `step4c` | `raw_data/hpa_liver/rna_tissue_consensus.tsv.zip` |
| Ensembl CDS FASTAs (×3) | **Auto-downloaded** by `step5a` | `genomes/Homo_sapiens.GRCh38.113.cds.all.fa.gz` etc. |

### Directory structure

```
BBB-Gene-Conservation/
├── raw_data/
│   ├── daneman_2010/       ← Daneman .xls files
│   ├── munji_2019/         ← Munji .xlsx file
│   ├── walchli_2024/       ← Seurat .rds.gz object
│   └── hpa_liver/          ← HPA .tsv (auto-downloaded)
├── genomes/                ← CDS FASTAs (auto-downloaded; not tracked by git)
├── processed/              ← All pipeline outputs (CSVs)
└── scripts/                ← All computation scripts
```

---

## Step 1 — Gene List

::: {.callout-note}
## Data flow
**Full script:** `scripts/step1_extract_BBB_genelist.R`
**Reads:** `raw_data/daneman_2010/Daneman2010_S3/S4/S5_*.xls`, `raw_data/munji_2019/Munji2019_S5_*.xlsx`
**Writes:** `processed/master_BBB_genelist.csv`
**Packages:** `readxl`, `dplyr`, `readr`, `homologene`
:::

### Why Daneman + Munji

Two foundational papers each profiled brain endothelial cells in mouse:

- **Daneman et al. 2010** — compared brain ECs to peripheral, liver, and lung ECs (supplementary tables S3, S4, S5). The S3 list is the strictest (genes enriched vs both liver AND lung).
- **Munji et al. 2019** — top CNS endothelial-enriched genes (supplementary table S5).

Mouse gene symbols from both papers were converted to human HGNC symbols using the `homologene` R package. Each gene was tagged by source paper and which Daneman filter it passed (`liver_and_lung`, `liver_only`, or `lung_only`).

### Why we expanded to S3+S4+S5

Initial work used only the strict S3 list (~545 genes). Canonical BBB genes like **SLC2A1** (GLUT1, the main brain glucose transporter) were missing — S3 demands enrichment vs liver AND lung simultaneously, and SLC2A1 is also expressed in liver endothelium. Including S4 and S5 with stringency tags captured these missing genes while preserving the ability to filter back to S3 for sensitivity analysis.

### Result

```{r master-list}
master <- read_csv("processed/master_BBB_genelist.csv", show_col_types = FALSE)

cat("Total genes in master list:", nrow(master), "\n\n")

table(master$Source) |>
  as.data.frame() |>
  rename(Source = Var1, Genes = Freq) |>
  kable()
```

```{r daneman-filter}
master |>
  filter(!is.na(Daneman_Filter)) |>
  count(Daneman_Filter) |>
  kable(caption = "Daneman stringency breakdown — `liver_and_lung` is the strictest")
```

---

## Step 2 — Wälchli

::: {.callout-note}
## Data flow
**Full script:** `scripts/step2_process_walchli_seurat.R`
**Reads:** `raw_data/walchli_2024/Walchli2024_..._SeuratObject.rds.gz`
**Writes:** `processed/Walchli2024_AverageExpression.csv`
**Packages:** `Seurat`, `dplyr`, `readr`
**Note:** Loading the Seurat object requires ~8 GB RAM and several minutes.
:::

The Wälchli 2024 dataset is a Seurat object containing 30,000+ human brain endothelial cells annotated by EC subtype (capillary, arteriole, venule, etc.).

### What we did

Used Seurat's `AverageExpression()` to collapse all cells into per-cell-type mean expression values, producing a gene × cell-type matrix.

The trickiest part: Seurat's default cell identities in this object were patient ages (15, 19 years), not cell types. Running `AverageExpression()` without overriding this would have averaged across patients, not cell types. Fix: `group.by = "ECclusters"`.

### Result

```{r walchli}
walchli <- read_csv("processed/Walchli2024_AverageExpression.csv",
                    show_col_types = FALSE)

cat("Wälchli expression matrix:", nrow(walchli), "genes x",
    ncol(walchli) - 1, "cell types\n\n")

cat("Cell types:\n")
cat(paste("  -", colnames(walchli)[-1], collapse = "\n"))
```

The **Mitochondrial** cluster was excluded from downstream analysis — cells with abnormally high mitochondrial gene content are typically dying or low-quality.

---

## Steps 3 & 3b — Mapping

::: {.callout-note}
## Data flow
**Full scripts:** `scripts/step3_map_ensembl_crossspecies.R` (superseded), `scripts/step3b_biomart_orthology_confidence.R`
**Reads:** `processed/master_BBB_genelist.csv` → `processed/master_BBB_genelist_validated.csv` + `processed/BBB_genes_crossspecies_ensembl.csv`
**Writes:** `processed/master_BBB_genelist_complete.csv`
**Packages:** `biomaRt`, `dplyr`, `readr`
**Note:** BioMart queries require internet and take 5–15 minutes. Results are cached.
:::

::: {.callout-warning}
## What went wrong — and why there is no Step 3a

The first attempt used the `homologene` R package — an offline NCBI orthologue database — to map all three species simultaneously. This worked for human → mouse but **returned no results for macaque**: HomoloGene does not include rhesus macaque (*Macaca mulatta*).

Step 3b re-ran everything via **Ensembl BioMart**, which supports all three species and additionally returns orthologue type and pre-computed % sequence identity. This is why the pipeline skips from Step 3 to Step 3b.
:::

### What we did

Three queries to Ensembl BioMart (each uses a different attribute page, so they cannot be combined):

1. Human gene symbol → human Ensembl gene ID (`ENSG...`)
2. Human Ensembl ID → mouse orthologue (`ENSMUSG...`)
3. Human Ensembl ID → macaque orthologue (`ENSMMUG...`)

Step 3b added `orthology_type` and `perc_id` per orthologue.

### Result

```{r ensembl-map}
combined <- read_csv("processed/master_BBB_genelist_complete.csv",
                     show_col_types = FALSE)

cat("Total rows in master table:", nrow(combined), "\n")
cat("  Human Ensembl IDs:   ", sum(!is.na(combined$HumanEnsembl)), "\n")
cat("  Mouse orthologues:   ", sum(!is.na(combined$MouseEnsembl)), "\n")
cat("  Macaque orthologues: ", sum(!is.na(combined$MacaqueEnsembl)), "\n")
```

```{r orthology-types}
combined |>
  filter(!is.na(Mouse_OrthoType)) |>
  count(Mouse_OrthoType) |>
  kable(caption = "Mouse orthologue type distribution")
```

The 1,724 rows (vs 1,526 input genes) reflects **one-to-many** orthologues — e.g. human ABCB1 maps to both mouse *Abcb1a* and *Abcb1b*. Both rows are kept and flagged.

---

## Steps 4b & 4c — Validation

::: {.callout-warning}
## What went wrong — and why Step 4 was removed

Step 4 downloaded full **GTF genome annotation files** (~1.5 GB per species, ×3 = ~4.5 GB total) with the intention of extracting CDS coordinates manually. After downloading all three we found that Ensembl already provides **pre-extracted CDS FASTA files** (~20 MB per species) — the same information at a fraction of the size. Step 4 was removed entirely, which is why the numbering jumps from Step 3b to Step 4b.
:::

### Step 4b — Human EC Validation

::: {.callout-note}
## Data flow
**Full script:** `scripts/step4b_human_validation.R`
**Reads:** `processed/master_BBB_genelist.csv`, `processed/Walchli2024_AverageExpression.csv` (+ Yang 2022 and Winkler 2022 gene lists, hardcoded)
**Writes:** `processed/master_BBB_genelist_validated.csv`
**Packages:** `dplyr`, `readr`
:::

The master BBB gene list was built from **mouse** studies. Step 4b checks which genes are also expressed in **human** brain endothelium using three independent datasets:

- **Wälchli 2024** — expressed > 0 in any EC cluster
- **Yang 2022** — 112-gene endothelial subtype marker list
- **Winkler 2022** — EC marker list from a large human brain vascular atlas

Each gene receives a count (0–3) of how many datasets confirm it.

```{r validation}
combined <- read_csv("processed/master_BBB_genelist_complete.csv",
                     show_col_types = FALSE)

combined |>
  count(Human_BBB_Datasets_N) |>
  rename(`# datasets confirming` = Human_BBB_Datasets_N, Genes = n) |>
  kable(caption = "Human EC validation score distribution")
```

74 genes have zero human EC evidence — they came purely from mouse data and are flagged in downstream analysis.

### Step 4c — Liver Control Gene Set

::: {.callout-note}
## Data flow
**Full script:** `scripts/step4c_liver_control_geneset.R`
**Reads:** `processed/master_BBB_genelist_complete.csv`, `raw_data/hpa_liver/rna_tissue_consensus.tsv` (auto-downloaded)
**Writes:** `processed/liver_control_genelist.csv`
**Packages:** `dplyr`, `readr`
:::

Liver was chosen as the control tissue — metabolically active and highly conserved, providing a meaningful non-BBB baseline. Filtered for genes with liver nTPM ≥ 10 that are NOT in the BBB master list.

```{r liver}
liver_ctl <- read_csv("processed/liver_control_genelist.csv", show_col_types = FALSE)
cat("Liver control set:", nrow(liver_ctl), "genes\n")
cat("Liver nTPM — median:", round(median(liver_ctl$Liver_nTPM), 1),
    " mean:", round(mean(liver_ctl$Liver_nTPM), 1), "\n")
```

Landmark liver genes confirmed present: ALB (198,523 nTPM), CYP3A4, APOB, TTR.

---

## Steps 5a-b — CDS

::: {.callout-note}
## Data flow
**Full scripts:** `scripts/step5a_download_cds_fastas.R`, `scripts/step5b_extract_cds_sequences.R`
**Step 5a** downloads Ensembl CDS FASTAs → `genomes/*.cds.all.fa.gz` (×3 species)
**Step 5b reads:** genome FASTAs + gene lists → **Writes:** `genomes/cds_BBB_*.fa` and `genomes/cds_liver_*.fa` (×3 species each)
**Packages:** `Biostrings`, `dplyr`, `readr`
**Note:** `genomes/` is excluded from git (too large). Run Step 5a to populate it before Step 5b.
:::

### What we did

Ensembl provides pre-extracted CDS FASTA files for every annotated transcript. For each gene with multiple transcripts (alternative splicing), we kept the **longest CDS** as a heuristic for the full open reading frame.

> **Alternative:** `scripts/step5b_alt_canonical_transcripts.R` uses **Ensembl Canonical / MANE Select** transcripts instead — one curated transcript per gene, jointly designated by NCBI + EMBL-EBI. More standard for publication-grade work. The FASTAs are already generated in `genomes/cds_canonical_*.fa`.

### Result

```{r cds-stats}
library(Biostrings)
bbb_human <- readDNAStringSet("genomes/cds_BBB_human.fa")
cat("Human BBB CDS sequences:", length(bbb_human), "\n")
cat("Length: median =", as.integer(median(width(bbb_human))), "bp\n")
cat("        mean   =", as.integer(mean(width(bbb_human))), "bp\n")
```

---

## Steps 5c-d — Scoring

::: {.callout-note}
## Data flow
**Full scripts:** `scripts/step5c_align_and_score.R` (BBB), `scripts/step5d_liver_scoring.R` (liver)
**Reads:** `genomes/cds_BBB_*.fa` or `genomes/cds_liver_*.fa`, `processed/master_BBB_genelist_complete.csv`
**Writes:** `processed/conservation_scores_BBB.csv`, `processed/conservation_scores_liver.csv`, `processed/conservation_scores_combined.csv`
**Packages:** `Biostrings`, `pwalign`, `seqinr`, `dplyr`, `readr`
**Note:** Step 5c takes 20–40 min (~1,500 alignments). Step 5d takes 60–90 min (~3,000 alignments).
:::

### Pairwise global alignment

For each gene we ran a **Needleman-Wunsch global alignment** between species pairs, then computed % identity and dN/dS:

```r
aln <- pwalign::pairwiseAlignment(
  seq1, seq2, type = "global",
  substitutionMatrix = nucleotideSubstitutionMatrix(match=1, mismatch=-1),
  gapOpening = -5, gapExtension = -2
)
pct_identity <- pwalign::pid(aln, type = "PID1")
# PID1 = matches / alignment_length (including gaps)
```

**Why global, not local?** These are orthologous genes — they should align end-to-end. Local alignment would only find the most similar segment, over-estimating similarity for genes that diverge at the edges.

### Why % identity AND dN/dS

- **% identity** — how similar are the letters, simple and interpretable
- **dN/dS** — how often do mutations *change the protein* (non-synonymous, dN) vs leave it unchanged (synonymous, dS). dN/dS << 1 means evolution is removing protein-changing mutations (purifying selection).

---

## Steps 5e-g — Statistics

::: {.callout-note}
## Data flow
**Full scripts:** `scripts/step5e_statistics.R`, `scripts/step5f_recompute_dnds.R`, `scripts/step5g_stats_v2.R`
**Step 5e** reads `conservation_scores_combined.csv` → writes `conservation_summary.txt`
**Step 5f** reads combined scores + CDS FASTAs → **writes** `conservation_scores_combined_v2.csv`
**Step 5g** reads v2 scores → writes `conservation_summary_v2.txt`
**Packages:** `Biostrings`, `pwalign`, `seqinr`, `dplyr`, `readr`
:::

### Step 5e: First-pass statistics

Wilcoxon rank-sum tests comparing BBB vs liver (non-parametric, because % identity is bounded and skewed). Stratified analyses by Daneman filter, validation score, and orthologue type.

### Step 5f: dN/dS recomputation

::: {.callout-warning}
## What went wrong — three separate dN/dS bugs

**Bug 1 — all NA values:** The `ape` package (loaded alongside `seqinr`) has its own `kaks()` function and silently overrode `seqinr::kaks()`. Fix: explicit namespace.

**Bug 2 — values inflated:** `seqinr::kaks()` assumes codon-aligned input — same length, same reading frame positions in both sequences. We were passing unaligned sequences, inflating the non-synonymous count. Especially large effect for human vs mouse (~80 million year divergence).

**Bug 3 — wrong translation function:** `dplyr`'s `translate()` (an SQL function) was silently overriding `Biostrings::translate()`. Fix: `Biostrings::translate(DNAString(seq), if.fuzzy.codon = "X")`.
:::

**The fix:** translate both sequences to amino acids → align proteins with BLOSUM62 → back-map codons using protein alignment → strip gap columns → pass codon-aligned pair to `seqinr::kaks()`:

```r
aa1 <- Biostrings::translate(DNAString(nt1), if.fuzzy.codon = "X")
aa2 <- Biostrings::translate(DNAString(nt2), if.fuzzy.codon = "X")
protein_aln <- pwalign::pairwiseAlignment(aa1, aa2, type = "global",
                 substitutionMatrix = "BLOSUM62",
                 gapOpening = 10, gapExtension = 4)
# back-map to codons, strip gaps, run seqinr::kaks()
# full implementation: scripts/step5f_recompute_dnds.R
```

### Step 5g: Multiple testing correction

Bonferroni and FDR (Benjamini-Hochberg) applied to the four primary BBB-vs-liver comparisons. Stratified tests are flagged as exploratory.

### Final results

```{r v2-results}
df <- read_csv("processed/conservation_scores_combined_v2.csv",
               show_col_types = FALSE)

bbb   <- df[df$Gene_Set == "BBB", ]
liver <- df[df$Gene_Set == "Liver_Control", ]

results <- data.frame(
  Comparison = c("Human vs Mouse — % identity",
                  "Human vs Macaque — % identity",
                  "Human vs Mouse — dN/dS (corrected)",
                  "Human vs Macaque — dN/dS (corrected)"),
  BBB_median = c(
    round(median(bbb$PctId_Human_Mouse,      na.rm = TRUE), 2),
    round(median(bbb$PctId_Human_Macaque,    na.rm = TRUE), 2),
    round(median(bbb$dNdS_Human_Mouse_v2,    na.rm = TRUE), 3),
    round(median(bbb$dNdS_Human_Macaque_v2,  na.rm = TRUE), 3)
  ),
  Liver_median = c(
    round(median(liver$PctId_Human_Mouse,      na.rm = TRUE), 2),
    round(median(liver$PctId_Human_Macaque,    na.rm = TRUE), 2),
    round(median(liver$dNdS_Human_Mouse_v2,    na.rm = TRUE), 3),
    round(median(liver$dNdS_Human_Macaque_v2,  na.rm = TRUE), 3)
  )
)
kable(results, caption = "BBB vs liver conservation — final corrected values")
```

```{r v2-pid-plot, fig.cap="Distribution of % nucleotide identity — BBB vs Liver"}
df_long <- df |>
  select(Gene_Set, PctId_Human_Mouse, PctId_Human_Macaque) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison",
                       values_to = "PctIdentity") |>
  filter(!is.na(PctIdentity)) |>
  mutate(Comparison = sub("PctId_", "", Comparison),
         Comparison = sub("_", " vs ", Comparison))

ggplot(df_long, aes(x = PctIdentity, fill = Gene_Set)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Comparison) +
  scale_fill_manual(values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4")) +
  labs(x = "% nucleotide identity", y = "Density",
       title = "BBB vs Liver: distribution of % identity", fill = "Gene set") +
  theme_minimal()
```

```{r v2-dnds-plot, fig.cap="Distribution of dN/dS (codon-aligned)"}
df_dnds <- df |>
  select(Gene_Set, dNdS_Human_Mouse_v2, dNdS_Human_Macaque_v2) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison", values_to = "dNdS") |>
  filter(!is.na(dNdS), dNdS < 2)

ggplot(df_dnds, aes(x = dNdS, fill = Gene_Set)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Comparison) +
  scale_fill_manual(values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4")) +
  labs(x = "dN/dS (codon-aligned)", y = "Density",
       title = "BBB vs Liver: distribution of dN/dS", fill = "Gene set") +
  theme_minimal()
```

```{r pid-vs-dnds-plot, fig.cap="% identity vs dN/dS — both metrics tell the same story"}
bbb_scatter <- bbb |>
  filter(!is.na(PctId_Human_Mouse), !is.na(dNdS_Human_Mouse_v2), dNdS_Human_Mouse_v2 < 2)

ggplot(bbb_scatter, aes(x = PctId_Human_Mouse, y = dNdS_Human_Mouse_v2)) +
  geom_point(alpha = 0.25, color = "#d62728", size = 1.2) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.8) +
  labs(x = "% nucleotide identity (Human vs Mouse)", y = "dN/dS (codon-aligned)",
       title = "% identity and dN/dS are correlated but capture different signal",
       subtitle = "BBB genes only — Human vs Mouse") +
  theme_minimal()
```

```{r purifying-selection-bar, fig.cap="Proportion of genes under strong purifying selection (dN/dS < 0.3)"}
purify_df <- df |>
  filter(!is.na(dNdS_Human_Mouse_v2) | !is.na(dNdS_Human_Macaque_v2)) |>
  group_by(Gene_Set) |>
  summarise(
    `Human vs Mouse`   = mean(dNdS_Human_Mouse_v2   < 0.3, na.rm = TRUE) * 100,
    `Human vs Macaque` = mean(dNdS_Human_Macaque_v2 < 0.3, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(cols = -Gene_Set, names_to = "Comparison", values_to = "Pct")

ggplot(purify_df, aes(x = Comparison, y = Pct, fill = Gene_Set)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("BBB" = "#d62728", "Liver_Control" = "#1f77b4")) +
  labs(x = NULL, y = "% of genes with dN/dS < 0.3",
       title = "Proportion of genes under strong purifying selection",
       subtitle = "dN/dS < 0.3 = conventional threshold for strong constraint",
       fill = "Gene set") +
  theme_minimal()
```

---

## Step 5h — Triangle

::: {.callout-note}
## Data flow
**Full script:** `scripts/step5h_mouse_vs_macaque.R`
**Reads:** `genomes/cds_BBB_mouse.fa`, `genomes/cds_BBB_macaque.fa`, full genome FASTAs, `processed/conservation_scores_combined_v2.csv`
**Writes:** `processed/conservation_scores_mouse_macaque.csv`
**Note:** ~1,500 alignments, ~30 minutes.
:::

Earlier steps used human as the reference. To complete the picture we computed the third side: mouse vs macaque directly.

```{r triangle}
mm     <- read_csv("processed/conservation_scores_mouse_macaque.csv",
                   show_col_types = FALSE)
mm_bbb <- mm[mm$Gene_Set == "BBB", ]

triangle <- data.frame(
  Comparison = c("Human  vs Mouse", "Human  vs Macaque", "Mouse  vs Macaque"),
  `Median % identity (BBB genes)` = c(
    round(median(bbb$PctId_Human_Mouse,      na.rm = TRUE), 1),
    round(median(bbb$PctId_Human_Macaque,    na.rm = TRUE), 1),
    round(median(mm_bbb$PctId_Mouse_Macaque, na.rm = TRUE), 1)
  ),
  check.names = FALSE
)
kable(triangle)
```

```{r translational-ladder-plot, fig.cap="Human vs Mouse and Human vs Macaque % identity per BBB gene"}
bbb_both <- bbb |>
  filter(!is.na(PctId_Human_Mouse), !is.na(PctId_Human_Macaque))

ggplot(bbb_both, aes(x = PctId_Human_Mouse, y = PctId_Human_Macaque)) +
  geom_point(alpha = 0.25, color = "#d62728", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey40", linewidth = 0.8) +
  annotate("text", x = 60, y = 98, label = "Macaque closer to human",
           color = "grey30", size = 3.5, hjust = 0) +
  annotate("text", x = 85, y = 72, label = "Mouse closer to human",
           color = "grey30", size = 3.5, hjust = 0) +
  labs(x = "Human vs Mouse % identity", y = "Human vs Macaque % identity",
       title = "For almost every BBB gene, macaque is more similar to human than mouse is",
       subtitle = "Points above the dashed diagonal = macaque is the closer model") +
  theme_minimal()
```

The complete evolutionary distance picture:

```
                Human
              /        \
          84.1%        96.8%
            /              \
        Mouse  ── 82.6% ── Macaque
```

---

## Results

### Headline findings

1. **Macaque is the closest model to human BBB.** 96.8% nucleotide identity at the coding-sequence level. For almost every gene, macaque sits above the diagonal — more similar to human than mouse.

2. **Mouse is more divergent (84.1%) but still highly conserved.** Mouse models remain biologically defensible, but ~16% sequence divergence is present. Where mouse and human BBB phenotypes differ clinically, this divergence is one possible explanation.

3. **BBB genes are under stronger purifying selection than liver genes** — most visibly in the human-macaque comparison (BBB dN/dS 0.152 vs Liver 0.196, p < 10⁻⁶ after Bonferroni correction). The signal is weaker in human-mouse because at ~80 million years, all genes have accumulated extensive change.

4. **Gene duplication produces dramatic divergence.** Paralogous mouse genes (e.g. *Abcb1a* + *Abcb1b* for human ABCB1) are only ~47% identical to human, vs ~85% for one-to-one orthologues. Effect size is large (rank-biserial r = 0.72).

5. **Conservation and human BBB expression are independent.** A gene's validation score (0–3 datasets) does not predict how conserved it is at the sequence level.

### Biological interpretation

Strong purifying selection on BBB genes makes biological sense. The BBB controls what enters and exits the brain — broken transporters or tight junctions would let toxins in or block essential nutrients. Evolution actively removes protein-changing mutations from these genes, more aggressively than for liver genes.

The fact that this signal is most visible in the human-macaque comparison — not human-mouse — reflects evolutionary timescale. Subtle differential selection is detectable when species are close enough that not all genes have saturated their substitutions. At ~80 million years for human-mouse, this signal is lost in the noise.

---

## Limitations

### Methodological caveats

- **Pairwise dN/dS, not phylogenetic.** A more rigorous analysis uses PAML `codeml` on a multi-species alignment with an explicit phylogenetic tree. Discuss with supervisor before publication.
- **Longest-transcript heuristic.** MANE Select / Ensembl Canonical (`step5b_alt_canonical_transcripts.R`) are more standard. The canonical FASTAs are already generated.
- **Multiple testing correction was applied to primary tests only.** Stratified subgroup analyses (by Daneman filter, validation score, paralogue type) are exploratory.

### Biological caveats

- **The starting gene list is enrichment-based.** CLDN5 (the canonical BBB tight junction protein) is absent because it failed the differential expression cutoff — it is expressed in all endothelia, not just brain.
- **Liver is one specific control choice.** A different control (brain non-BBB genes, genes matched by protein domain) could give a different picture.
- **Coding sequence only.** Regulatory regions (promoters, enhancers, UTRs) — where most cross-species expression-level differences live — were deliberately excluded.

### Future extensions

1. **PAML codeml** for proper phylogenetic dN/dS.
2. **Promoter sequence comparison** (~2 kb upstream of TSS) to test regulatory divergence.
3. **MANE Select transcripts** (already scripted — swap in `genomes/cds_canonical_*.fa`).
4. **Brain non-BBB genes** as a complementary control set.
5. **Giger macaque dataset** — are the most conserved BBB genes also the most highly expressed in macaque endothelium?
6. **Functional category breakdown** — are tight junction proteins more conserved than transporters? Are drug efflux pumps the most protected class?

---

## References

### Datasets

- Daneman R, et al. (2010). The mouse blood-brain barrier transcriptome. *PLOS ONE* 5(10): e13741.
- Munji RN, et al. (2019). Profiling the mouse brain endothelial transcriptome in health and disease. *Nature Neuroscience* 22(11): 1892-1902.
- Wälchli T, et al. (2024). Single-cell atlas of the human brain vasculature. *Nature* 614: 945-954.
- Yang AC, et al. (2022). A human brain vascular atlas reveals diverse mediators of Alzheimer's risk. *Nature* 603: 885-892.
- Winkler EA, et al. (2022). A single-cell atlas of the normal and malformed human brain vasculature. *Science* 375(6584): eabi7377.

### Tools

- Ensembl release 113. https://www.ensembl.org
- BioMart (`biomaRt` R package).
- Human Protein Atlas. https://www.proteinatlas.org
- Bioconductor: `Biostrings`, `pwalign`, `seqinr`.

### AI Transparency

This pipeline was developed by Yara Ismail with the assistance of Claude (Anthropic) for R script writing and debugging. All scientific decisions, interpretation, and research design were made by the author under supervision of Dr. Rachel Clelland.

---

:::

*Render with `quarto render docs/BBB_Conservation_Tutorial.qmd` from the project root.*
